---
title: "Exam1 Part D. Clustering practice"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(cluster)
library(dplyr)
library(magrittr)
library(ggplot2)
library(plotly)
library(data.table)
library(caret)
if (!require("ggbiplot")) install.packages("ggbiplot")
library(tidyr)
if (!require("mltools")) install.packages("mltools")
if (!require("cluster")) install.packages("cluster")
if (!require("factoextra")) install.packages("factoextra")
library(cluster)
library(factoextra)

current_date <- as.Date("2014-07-01")
```

# Notes on Part D submission

Complete this Rmarkdown document, knit it to html or pdf and submit it along with this Rmd file. Write


# Problem Statement

An advertisement division of large club store needs to perform customer analysis 
the store customers in order to create a segmentation for more targeted marketing campaign 

Your task is to identify similar customers and characterize them (at least some of them). 
In other word perform clustering and identify customers segmentation.

```
Colomns description:
People
  ID: Customer's unique identifier
  Year_Birth: Customer's birth year
  Education: Customer's education level
  Marital_Status: Customer's marital status
  Income: Customer's yearly household income
  Kidhome: Number of children in customer's household
  Teenhome: Number of teenagers in customer's household
  Dt_Customer: Date of customer's enrollment with the company
  Recency: Number of days since customer's last purchase
  Complain: 1 if the customer complained in the last 2 years, 0 otherwise

Products

  MntWines: Amount spent on wine in last 2 years
  MntFruits: Amount spent on fruits in last 2 years
  MntMeatProducts: Amount spent on meat in last 2 years
  MntFishProducts: Amount spent on fish in last 2 years
  MntSweetProducts: Amount spent on sweets in last 2 years
  MntGoldProds: Amount spent on gold in last 2 years

Place
  NumWebPurchases: Number of purchases made through the company’s website
  NumStorePurchases: Number of purchases made directly in stores
```

Assume that data was current on 2014-07-01

# 1. Read Dataset and Data Conversion to Proper Data Format
# Convert Dt_customer to numeric dates
# Convert Dt_customer to no_days
Read "m_marketing_campaign.csv" using `data.table::fread` command, examine the data.

> `fread` function of `data.table` read cvs real fast

```{r}
# fread m_marketing_campaign.csv and save it as df
df <- data.table::fread('m_marketing_campaign.csv')

# Checking how the data looks like
head(df)

# Basic stats of the dataframe
summary(df)
```



```{r}
# Convert Year_Birth to Age (assume that current date is 2014-07-01)

# To convert Year_Birth to Age, generating random days and months to attach to the year.
# Assuming days varies from 1 - 28 and month varies from 1 - 12

# Random dates uniformly chosed between 1- 28
Day = floor(runif(length(df$Year_Birth), min=1, max=28))

# Random months uniformly chosen between 1 - 12
Month = floor(runif(length(df$Year_Birth), min=1, max=12))
Year = df$Year_Birth

# Created another data frame using the year, month and day
dyb = data.frame(Year, Month, Day)
# Combined all to create dates
dyb$Year_birth = as.Date(with(dyb, paste(Year, Month, Day, sep='-')), "%Y-%m-%d")
# Added age columns using the birthdates created for all the rows
df$Age = as.numeric(difftime(current_date, dyb$Year_birth, units='weeks'))/52.25

# Dt_Customer is a date (it is still character), convert it to membership days (name it MembershipDays)
# hint: note European date format, use as.Date with proper format argument

# Calculated membership days based on the Dt_Customer
df$MembershipDays = as.numeric(difftime(current_date, as.Date(df$Dt_Customer, "%d-%m-%Y")))

```

```{r}
# Summarize Education column (use table function)
table(df$Education)

# Lets treat Education column as ordinal categories and use simple levels for distance calculations
# Assuming following order of degrees:
#    HighSchool, Associate, Bachelor, Master, PhD
# factorize Education column (hint: use factor function with above levels)
df$Education = factor(df$Education, order=TRUE, levels = c("HighSchool", "Associate", "Bachelor", "Master", "PhD"))

```

```{r}
# Summarize Education column (use table function)
table(df$Marital_Status)

# Lets convert single Marital_Status categories for 5 separate binary categories 
# Divorced, Married, Single, Together and Widow, the value will be 1 if customer 
# is in that category and 0 if customer is not
# hint: use dummyVars from caret package, model.matrix or simple comparison (there are only 5 groups)

# One Hot encoding of Marital_Status
dum = dummyVars(" ~ Marital_Status", df)
ms_oh <- data.frame(predict(dum, newdata = df))
df = dplyr::bind_cols(df, ms_oh)

```

```{r}
# lets remove columns which we will no longer use:
# remove ID, Year_Birth, Dt_Customer, Marital_Status
# and save it as df_sel

# Removing ID, Year_Birth, Dt_Customer, Marital_Status from the dataframe
df_sel = select(df, -1:-2, -4, -8)

# Convert Education to integers 
# hint: use as.integer function, if you use factor function earlier 
# properly then HighSchool will be 1, Associate will be 2 and so on)
df_sel$Education = as.numeric(df$Education)

```


```{r}
# lets scale
# run scale function on df_sel and save it as df_scale
# that will be our scaled values which we will use for analysis

df_scale = scale(df_sel)
```

(5 points)

# 2. Run PCA

```{r}
# Run PCA on df_scale, make biplot and scree plot/percentage variance explained plot
# save as pc_out, we will use pc_out$x[,1] and pc_out$x[,2] later for plotting
pc_out = prcomp(df_scale, scale = TRUE)
#plot(pc_out$sdev)
if (!require("devtools")) install.packages("devtools")
library("devtools")
install_github("vqv/ggbiplot")
require(ggbiplot)
ggbiplot(pc_out)

# Biplot using first two principal components
biplot = ggbiplot(pcobj = pc_out,
                  choices = c(1, 2),
                  var.axes = FALSE)

biplot = biplot + labs(title = "PCA biplot using PC1 / PC2") + theme_minimal()

print(biplot)

```

# Scree plot
```{r}

# Variance explained by each principal components
var_explained_df <- data.frame(PC= c(1:21),
                               var_explained=(pc_out$sdev)^2/sum((pc_out$sdev)^2))

var_explained_df %>%
  ggplot(aes(x=PC,y=var_explained, group=1))+
  geom_point(size=4)+
  geom_line()+
  labs(title="Scree plot: PCA on scaled data")
```

`Question: (5 points)`
Comment on observation (any visible distinct clusters?):

As per the biplot of PC1 and PC2, there is no clear segregation of clusters. 
One reason for this can be that there is no cluster in the data. Another reason 
can be that, since the visualization is only using two principal components, we 
are a not able to see clusters of 21-dimensional subspace in 2-D.

# 3. Cluster with K-Means
## 3.1 Selecting Number of Clusters

Run K-Means for a range of k-s select one to use later 
(save it as k_kmeans)

```{r}

km_out_list <- lapply(1:10, function(k) list(
  k=k,
  km_out=kmeans(df_scale, k, nstart = 25)))

km_results <- data.frame(
  k=sapply(km_out_list, function(k) k$k),
  totss=sapply(km_out_list, function(k) k$km_out$totss),
  tot_withinss=sapply(km_out_list, function(k) k$km_out$tot.withinss)
  )
km_results
plot_ly(km_results,x=~k,y=~tot_withinss) %>% add_markers() %>% add_paths()



# Gap statistics

gap_kmeans <- clusGap(df_scale, kmeans, nstart = 10, K.max = 20, B = 100)

plot(gap_kmeans, main = "Gap Statistic: kmeans")

```
```


```

```
`Question: (5 points)`
Which k did you choose and why?

# No clear elbow shown in the graph using entire data. While there is somewhat sharp decrease from 1 to 2,
# suggesting there might be two clusters (not sharp enough to be called en elbow). I see that there is a
# constant decrease (almost linear), suggesting there might be a higher number of clusters.

# Similar observation using gap-statistics. There is an increase from the 1 to 2
# but that is not the peak. Constant increase in the cluster, and a peak is observed
# at 10 clusters.

# As per the gap statistics, the peak is increasing as the number of clusters are increasing
# We also observed an increase on 2 clusters. This might suggest that there are more number of clusters
# in the data. If we gap statistics is run for 10 clusters, the peak is at 10, if it is run for
# 20 clusters, the peak is at 20. Even the elbow plot somewaht suggests that there 
# might be a higher number of clusters even though there is a drop at 2 clusters.

# I would be choosing 2 as the value of k because both gap statistics and elbow method suggest
# that there might be 2 clusters. This can be an over simplistic way of categorising the data
# but it is easier to humanly explain based on the amount of data and information present.
# Also, the number of observations would be substantial in both the clusters. If we choose
# a higher number of clusters, say, 10 or 20, it will be harder to explain what differentiates
# them in English and the number of data points in the clusters might vary a lot.

## 3.2 Clusters Visulalization

Make k-Means clusters with selected k_kmeans (store result as km_out).
Plot your k_kmeans clusters on biplot (just PC1 vs PC2) by coloring points by their cluster id.

```{r}
k_kmeans = 2
#pca_transform = as.data.frame(-pc_out$x[,1:2])
km_out <- kmeans(df_scale, k_kmeans, nstart = 20)

plot(pc_out$x[,1:2], col = (km_out$cluster + 1),
    main = "K-Means Clustering Results with K = 2",
    xlab = "PC1", ylab = "PC2", pch = 20, cex = 1)

#, "#449933", "#888888", "#0D69D0", "#999999", "#000000", "#420420"), 
fviz_cluster(km_out, data = df_scale,
             palette = c("#2E9FDF", "#00AFBB", "#449933"), 
             geom = "point",
             ellipse.type = "convex", 
             ggtheme = theme_bw()
             )

# km_out = k_kmeans$cluster
# plot(km_out)

# biplot = biplot + labs(title = "K-means biplot for PC1 / PC2") + theme_minimal()
```
`Question: (5 points)`
Do you see any grouping? Comment on you observation.

Based on the plot of PC1 and PC2, we can see that there are two centroids on the left
and the right which is diving the data into two clusters. Since, the number of clusters
are only 2 and the graph is color coded, the clusters can be seen now but there is still
no cluster visible with naked eyes. If we had more number of clusters, say 8 or 15,
the data points of different clusters would have been superimposed on one another making
it more difficult to differentiate the data among clusters.

Therefore, 2 clusters are clearly diving the data between two groups


## 3.3 Characterizing Cluster

Perform descriptive statistics analysis on obtained cluster. Based on that does one or more group have a distinct characteristics?
(5 points)
Hint: add cluster column to original df dataframe

```{r}

df_sel_new = df_sel
df_sel_new$Cluster = km_out$cluster


tapply(df_sel_new$Income, df_sel_new$Cluster, summary)
tapply(df_sel_new$MntFruits, df_sel_new$Cluster, summary)
tapply(df_sel_new$MntFishProducts, df_sel_new$Cluster, summary)

```
We can see in the summary of these columns based on two clusters that
customer falling in cluster 1 is falling under relatively less income
and less purchasing power group. The mean of income is almost half of
the mean of income of customers falling in cluster 2.
Even in the case of purchasing power, cluster 1 is spending much less
on fruits and meat than cluster two. The median of cluster 1 spending
on fruit is 3.0 while it is 35.0 for cluster 2. Similarly, the median
of cluster 1 spending on meat is a mere 4.0 while, it is 58.0 for
cluster 2.


# 4. Cluster with Hierarchical Clustering


Perform clustering with Hierarchical method.
Try complete, single and average linkage.
Plot dendagram, based on it choose linkage and number of clusters, if possible, explain your
choice.

(5 points)

```{r}
df_dist <- dist(df_scale)
hc.complete <- hclust(dist(df_scale), method = "complete")
hc.average <- hclust(dist(df_scale), method = "average")
hc.single <- hclust(dist(df_scale), method = "single")

#Please visualize the cluster one by one for each of the hierarchical
#clusters. Reducing the cex to 0.00001 to make the dendrogram less messy

par(mfrow = c(1, 1))
plot(hc.complete, main = "Complete Linkage",
    xlab = "", sub = "", cex = .00001)
    
    
plot(hc.average, main = "Average Linkage",
    xlab = "", sub = "", cex = .00001)
    
    
plot(hc.single, main = "Single Linkage",
    xlab = "", sub = "", cex = .00001)

uh_cut <- cutree(hc.average, 16)
plot(hc.average, cex = 0.00001)
rect.hclust(hc.average, k = 16, border = 5:10)


uh_cut <- cutree(hc.complete, 16)
plot(hc.complete, cex = 0.00001)
rect.hclust(hc.complete, k = 16, border = 5:10)

```
Hierarchichal clustering with complete linkage is showing different clusters with a very high number of cluster.
With 16 cuts, we are able to differentiate a few clusters in the dendogram but with a small number of clusters,
the we observe that there is one big cluster and other clusters are having very small number of observations,
more like outliers.
